Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
i hate shortcut notation

Radiative cooling can, however, be stabilized by the other source of (stable) wind cooling, such as the last term $\propto c_{s}^{2} \propto T$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$.  To assess the conditions for thermal stability we compare the nominal rate of wind heating to the rate of radiative cooling $\dot{q}_{\rm cool} =\Lambda(T) n^2$ at the stagnation radius,
\begin{align}
\left.\frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhostarrs}) and
(\ref{eq:rhors}).  Here $n = \rho/\mu m_p$ and we have approximated the cooling function $\Lambda(T)$ assuming solar metallicity gas using piece-wise power law (\citealt{Draine:2011a}, Fig. 34.1):
\begin{align}
\Lambda(T)\simeq
  \begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}&, T > 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}&, T < 2 \times 10^7 \text{K}     
  \end{cases}
  \label{eq:cooling2}
\end{align}
Note that at the stagnation radius equation (\ref{eq:cooling2}) is equivalent to the ratio of the cooling to free-fall timescale, i.e.
\begin{align}
\left. \frac{\tcool}{\tff}\right|_{\rs}&= \frac{9}{5} \left. \frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs},
\end{align}
where we have used equations~(\eqref{eq:Tanalytic}) and (\eqref{eq:rhors2}).

There thus exists a critical minimum wind velocity required for thermal stability at the stagnation radius
\begin{align}
v_{\rm crit} \simeq
  \begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
In $\S\ref{sec:numerical}$ we will show that $v_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.

If cooling becomes important in the flow, it may lead to thermal
instabilities.  There are two kinds of ``instabilities", global and
local.  The global instability is simply that, if cooling becomes very
rapid, the gas loses thermal pressure and starts to inflow faster,
i.e. the entire CNM accretes (e.g. {\bf AG citation missing here.}).  In
practice this should could manifest as a large increase in the
stagnation radius, due an effective {\it decrease} in the effective
heating rate.  This global ``instability" is not a true instability
insofar as the gas does not necessarily clump locally into low-T
clouds.

However, \citet{McCourt+12} argue that even if cooling is exactly
balanced by heating everywhere, there is still a local thermal
instability that operates to turn hot gas into cold clouds.  However,
this local instability only causes a significant local drop-out of
mass from the hot medium (to low-T clouds), if the cooling timescale,
\begin{align}
\frac{\tcool}{\tff} &\equiv \frac{3\rho k T/2 \mu \mp}{\dot{q}_{\rm cool}}
\end{align} 
is locally shorter than the approximately 10 times the free-fall
timescale $t_{\rm ff}$.

@@

Our results show that the location of the stagnation rate and resulting accretion rate, is very sensitive to the assumed heating
rate $\propto v_{w,0}^{2}$.  Physically, we expect that the heating rate will vary from galaxy to galaxy.. For example, SNe Ia represent a
potentialy important source of heating, but since they are rare they
will only be relevant in a time-averaged sense if the supernova rate
exceeds the inflow time of the gas near the stagnation point.

In this section we discuss potential sources of heating and provide estimates of the value of $\vwO$ corresponding to each.  Before proceeding, note that $\vwO$ is related to the heating rate per volume $e$ according to 
\begin{align}
  &\frac{q \vwO^2}{2}=e \,\,\,\Rightarrow\,\,\,\vwO = \sqrt{\frac{2 t_h e}{\eta \rhostar}}
  \label{eq:vw_eff}
\end{align}

@@

\subsubsection{Millisecond Pulsars} The spin down power of millisecond
pulsars (MSPs) could in principle be an important heating source.  If
the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each
contributes an average spin-down luminosity $\bar{L}_{\rm sd}$, then the
resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm
  msp}\epsilon$ corresponds to an effective heating rate (eq.~
[\ref{eq:vw_eff}])

\begin{align}
  v_{\rm w,0}^{\rm p} &=&\left({\frac{2\epsilon t_h n_{\rm msp} \bar{L}_{\rm sd}}{\eta}}\right)^{1/2} \nonumber \\
  &\approx& 90\epsilon^{1/2}\left(\frac{\bar{L}_{\rm
        sd}}{10^{34}\,{\rm erg\,s^{-1}}}\right)^{1/2}
  \eta_{0.02}^{-1/2}\,{\rm km\,s^{-1}},
  \end{align}

  where $\epsilon< 1$ is the thermalization efficiency of the wind.
  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass
  $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to
  $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$, a number we
  assume in the numerical estimate.  Using the ATNF radio pulsar
  catalog (\citealt{Manchester+05}), we estimate the average spin-down
  luminosity of millisecond pulsars in the field (excluding those in
  globular clusters) to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg
  s$^{-1}$ {\bf For eta=0.02 we cannot really
    say that vw from msps is negligible}

  If we assume a higher spin-down luminosity $L_{\rm sd}\simeq
  10^{35}$ ergs s$^{-1}$ characteristic of some Fermi pulsars, then we
  find that $v_{\rm w,0}^{\rm p} \simeq 270\epsilon^{1/2}
  (\eta/0.1)^{-1/2}$ km s$^{-1}$, in which case MSP heating can be
  comparable to main sequence wind heating for $\eta = 0.1$, $\epsilon
  = 1$.  However, much lower thermalization efficiencies of $\epsilon
  \lesssim 0.1$ are inferred based on modeling the insterstellar
  mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}.
  Also, the binaries giving rise to MSPs may be disocciated by stellar
  interactions in the dense nuclear cluster, reducing their numbers as
  compared to the field population.
\subsubsection{Type Ia Supernovae}

Energy injected by Type Ia supernovae (SNe Ia) represents a
potentially important source of heating, which unlike core collapse
SNe is present even in an evolved stellar population.  Assuming that
each Ia injects a thermal energy $E_{\rm Ia}$ into the interstellar
medium, and that the SNe rate (per stellar mass) is given by $R_{\rm
  Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm
  Ia}$ produces an effective wind heating parameter
(equation~[\ref{eq:vw_eff}])
\begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the CNM.

>>>>

Radiative cooling can, however, be stabilized by the other source of (stable) wind cooling, such as the last term $\propto c_{s}^{2} \propto T$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$.  To assess the conditions for thermal stability we compare the nominal rate of wind heating to the rate of radiative cooling $\dot{q}_{\rm cool} =\Lambda(T) n^2$ at the stagnation radius,
\begin{align}
\left.\frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhostarrs}) and
(\ref{eq:rhors}).  Here $n = \rho/\mu m_p$ and we have approximated the cooling function $\Lambda(T)$ assuming solar metallicity gas using piece-wise power law (\citealt{Draine:2011a}, Fig. 34.1):
\begin{align}
\Lambda(T)\simeq
  \begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}&, T > 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}&, T < 2 \times 10^7 \text{K}     
  \end{cases}
  \label{eq:cooling2}
\end{align}
There thus exists a critical minimum wind velocity required for thermal stability at the stagnation radius
\begin{align}
v_{\rm crit} \simeq
  \begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
In $\S\ref{sec:numerical}$ we will show that $v_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.  Our estimate in equation (\ref{eq:cooling2}) assumes the existence of a stagnation radius according to equation (\ref{eq:rs}), the validity of which requires $v_{\rm w} \gtrsim \sigma \sim v_{\rm crit}$.

If cooling becomes important in the flow ($v_{w} < v_{\rm crit}$), it may lead to thermal instabilities.  There are two kinds of ``instabilities", global and local.  The global instability is simply that, if cooling becomes very rapid, the gas loses thermal pressure and starts to inflow faster, i.e. the entire CNM accretes.  In
practice this should could manifest as a large increase in the stagnation radius, due an effective {\it decrease} in the effective heating rate.  This global ``instability" is not a true instability insofar as the gas does not necessarily clump locally into low-T clouds.

However, \citet{McCourt+12} argue that even if cooling is exactly balanced by heating everywhere, there is still a local thermal instability that operates to turn hot gas into cold clouds.  However, this local instability only causes a significant local drop-out of mass from the hot medium (to low-T clouds), if the cooling timescale,
\begin{align}
\frac{\tcool}{\tff} &\equiv \frac{3\rho k T/2 \mu \mp}{\dot{q}_{\rm cool}} \approx \frac{9}{5} \left. \frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs}
\end{align} 
is locally shorter than the approximately 10 times the free-fall
timescale $t_{\rm ff}$, where the last equality is evaluated at the stagnation radius and makes use of equations~(\ref{eq:Tanalytic}) and (\ref{eq:rhors2}).  We thus see that the condition $v_w \gtrsim v_{\rm crit}$ is also the condition for local instability, although the prefactor may be slightly different.

@@

Important properties such as the location of the stagnation radius, the SMBH accretion rate, and the possibility of thermal instability depend sensitively on the assumed heating rate $\propto v_{w,0}^{2}$  This section discusses individual heating sources, providing estimates of the value of $\vwO$ corresponding to each.  Before proceeding, note that $\vwO$ is related to the volumetric heating rate $e$ according to 
\begin{align}
  &\frac{q \vwO^2}{2}=e \,\,\,\Rightarrow\,\,\,\vwO = \sqrt{\frac{2 t_h e}{\eta \rhostar}}
  \label{eq:vw_eff}
\end{align}

@@

\subsubsection{Millisecond Pulsars} Energy injection due to the spin-down of millisecond pulsars (MSPs) is a potentially important heating source.  If the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each contributes an average spin-down luminosity $\bar{L}_{\rm sd}$, then the resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm
  msp}\epsilon$ corresponds to an effective heating rate (eq.~ [\ref{eq:vw_eff}])
\begin{eqnarray}
  v_{\rm w,0}^{\rm MSP} &=&\left({\frac{2\epsilon t_h n_{\rm msp} \bar{L}_{\rm sd}}{\eta}}\right)^{1/2} \nonumber \\
  &\approx& 90\epsilon^{1/2}\left(\frac{\bar{L}_{\rm
        sd}}{10^{34}\,{\rm erg\,s^{-1}}}\right)^{1/2}
  \eta_{0.02}^{-1/2}\,{\rm km\,s^{-1}},
  \end{eqnarray}
where $\epsilon< 1$ is the thermalization efficiency of the wind.  The above estimate adopts a pulsar density $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$ motivated by the $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}).

From the ATNF radio pulsar catalog (\citealt{Manchester+05}), we estimate that the average spin-down luminosity of millisecond pulsars in the field (excluding those in globular clusters) to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$, in which case $v_{\rm w,0} \lesssim 100$ km s$^{-1}$ is unimportant even for $\epsilon = 1$.  For higher spin-down luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of some Fermi-detected pulsars, then the higher value of $v_{\rm w,0} \lesssim 300$ km s$^{-1}$ makes MSP heating in principle be important for $\epsilon = 1$.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}.  Also, the binaries giving rise to MSPs may be disocciated by stellar interactions in the dense nuclear cluster, reducing their numbers as compared to the field population.  We thus conclude that MSPs are unlikely to be an important source of heating.

\subsubsection{Type Ia Supernovae}

Energy injected by Type Ia supernovae (SNe Ia) represents a
potentially important source of heating, which unlike core collapse
SNe is present even in an evolved stellar population.  Assuming that
each Ia injects a thermal energy $E_{\rm Ia}$ into the interstellar
medium, and that the SNe rate (per stellar mass) is given by $R_{\rm
  Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm
  Ia}$ produces an effective wind heating parameter
(equation~[\ref{eq:vw_eff}])
\begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.
